Ore Polynomials in Sage

### 2013-06-07 paper/main.tex
paper, section 5: guessing

\documentclass{amsart}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{bbm}
\usepackage{a4wide}
\usepackage{sagetex}

@@

\begin{sageexample}
  sage: load("ore_algebra.py");  #### TO BE FIXED ####
  sage: R.<x> = PolynomialRing(ZZ); A.<Dx> = OreAlgebra(R)
  sage: A
  sage: A.random_element()
  sage: Dx*x
  sage: B.<Sx> = OreAlgebra(R)
  sage: B
  sage: Sx*x
\end{sageexample}

@@

\section{Further Examples}

Some longer session showing how to solve a meaningful problem by using several features of the code. Ideas?

\subsection{}

\subsection{}

>>>>

\documentclass{amsart}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{bbm}
\usepackage{a4wide}
\usepackage{sagetex}
\usepackage{tikz}

@@

\begin{sageexample}
  sage: load("ore_algebra.py", "guessing.py");  #### TO BE FIXED ####
  sage: R.<x> = PolynomialRing(ZZ); A.<Dx> = OreAlgebra(R)
  sage: A
  sage: A.random_element()
  sage: Dx*x
  sage: B.<Sx> = OreAlgebra(R)
  sage: B
  sage: Sx*x
\end{sageexample}

@@

[[CITE RUBEY.]]

Guessing is one of the most popular features of software packages like gfun.  It
is a technique for detecting patterns in sequences of numbers, which is often
used in combinatorics. The basic idea is simple. If we are given a finite array
of numbers, though of as the first terms of an infinite sequence, we may want to
know whether this sequence satisfies a recurrence. The algorithm behind a
guessing engine searches for small equations matching the given data.
Generically, no such equations exist, so if some are found, it is fair to
``guess'' that they are in fact valid equations for the whole infinite sequence.

We provide a guessing function which takes as input a list of terms and an Ore
algebra, and returns as output an operator which matches the given data and
which, in some measure, would be unlikely to exist for random data.

\begin{sageexample}
  sage: data = [ 0, 1, 1, 2, 3, 5, 8, 13, 21, 34, 55 ]
  sage: L = guess(data, OreAlgebra(ZZ['n'], 'Sn'))
  sage: L
  sage: L(data)
  sage: M = guess(data, OreAlgebra(ZZ['x'], 'Dx'))
  sage: M
  sage: M(x/(1-x-x^2))
\end{sageexample}

If an algebra of differential operators is supplied as second argument, the data
is understood as the first few coefficients of a power series. The output
operator is expected to have this power series as solution.

It can happen that the procedure is unable to find an operator matching the 
given data. In this case, an exception is raised. There are two possible 
explanations for such an event. Either the sequence in question does not 
satisfy any equations, or it does but the equations are so big that more
data is needed to detect them.

A number of options are available for governing the search for relations.  In
order to explain them, we first need to give some details on the underlying
algorithms. For simplicity of language, we restrict here to the case of
recurrence operators. The situation for differential operators is very similar.

For the most typical situations, there are two important hyperbolas. One
describes the region in the $(r,d)$-plane consisting of all points for which
there exists an operator of order~$r$ and degree~$d$ truly satisfied by the
sequence in question. (See~\cite{..} for an explanation why the boundary of this
region is usually a hyperbola.) The second describes the region of all points
$(r,d)$ for which an operator of order~$r$ and degree~$d$ can be detected when
$N$ terms are provided as input. This region is determined by the requirement
$(r+1)(d+2)<N$.

The method tests a sequence of points $(r_1,d_1)$, $(r_2,d_2)$, \dots\ right
below this second parabola. Success at a point $(r_i,d_i)$ means that some
evidence for an operator of order $\leq r_i$ and degree $\leq d_i$ has been found. 
This operator however is not explicitly computed. Instead, the method uses the
partial information found about this operator to calculate an operator which with 
high probability is the minimal order operator satisfied by the sequence in question. 
This operator is usually more interesting than the one at $(r_i,d_i)$, and its 
computation is usually more efficient.

Using the option \verb|path|, the user can specify a list of points $(r_i,d_i)$
which should be used instead of the standard path. By setting the options \verb|min_degree|,
\verb|max_degree|, \verb|min_order|, \verb|max_order|, all points $(r,d)$ of the path
are discarded for which $r$ or $d$ is not within the specified bounds. These options
can be used to accelerate the search in situations where the user has some knowledge
(or intuition) about the size of the expected equations.

\hangindent=-6.3cm\hangafter=-14\leavevmode
\smash{\raise-5.3cm\rlap{\kern\hsize\kern-5.1cm\begin{tikzpicture}[scale=.45]
  \draw[->] (10.5,0) node {$r$} (-.2,0) -- (10, 0) ;
  \draw[->] (0,10.5) node {$d$} (0,-.2) -- (0, 10) ;
  \draw (1.2, 3.5) -- (1.4, 4) (1.2, 3) -- (1.6, 4) (1.2, 2.5) -- (1.8, 4) (1.2, 2) -- (2, 4)
        (1.4, 2) -- (2.2, 4) (1.6, 2) -- (2.4, 4) (1.8, 2) -- (2.6, 4) (2, 2) -- (2.8, 4) 
        (2.2, 2) -- (3, 4) (2.4, 2) -- (3.2, 4) (2.6, 2) -- (3.4, 4) (2.8,2)--(3.6,4) (3,2)--(3.8,4)
        (3.2, 2) -- (4, 4) (3.4, 2) -- (4.2, 4) (3.6, 2) -- (4.4, 4) (3.8,2)--(4.6,4) (4,2)--(4.8,4)
        (4.2, 2) -- (5, 4) (4.4, 2) -- (5.2, 4) (4.6, 2) -- (5.4, 4) (4.8,2)--(5.6,4) (5,2)--(5.8,4); 
  \fill[lightgray] (.75, 10) .. controls (1.5,2.5) and (2.5,1.5) .. (10, .75) -- (10, 10) -- (.75, 10);
  \draw[thick] (2, 10) .. controls (2, 1.2) and (2, 1.2) .. (10, 1.2); 
  \draw (-.2, 2) node {\vbox{\llap{\footnotesize min }\kern-3pt\llap{\footnotesize degree }}} -- (10, 2) 
        (-.2, 4) node {\vbox{\llap{\footnotesize max }\kern-3pt\llap{\footnotesize degree }}} -- (10, 4);
  \draw (1.2, -.2) node {\hbox to0pt{\rule{0pt}{2em}\hss\footnotesize min order\hss}} -- (1.2, 10)
        (5, -.2) node {\hbox to0pt{\rule{0pt}{2em}\hss\footnotesize max order\hss}} -- (5, 10);
  \fill (2.5, 3) circle(4pt) node[right] {\rule[-1em]{0pt}{0pt}$A$} (2, 9) circle(4pt) node[right] {$B$} ;
\end{tikzpicture}}}%
The figure on the right illustrates the typical situation for guessing problems
that are not too small and not too artificial. The light gray region indicates
the area which is not accessible with the given amount of data. Only the points
$(r,d)$ below it can be tested for an operator of order~$r$ and degree~$d$ that
fits to the given data. Let's assume that operators exist on and above the solid
black hyperbola. The user will usually not know this curve in advance but may
have some expectations about it and can restrict the search accordingly, for
example to the dashed area shown in the figure. The method will detect the
existence of an operator, say at point~$A$, and construct from the information
gained at this point an operator of minimal possible order, which may correspond
to point~$B$.  This operator is returned as output. Note that the degree of the
output may exceed the value of \verb|max_degree|, and its order may be smaller
than $\verb|min_order|$:

\begin{sageexample}
sage: data = [(n+1)^10*2^n + 3^n for n in xrange(200)]
sage: L = guess(data, OreAlgebra(ZZ['n'],'Sn'), min_order=3, max_degree=5)
sage: L.order(), L.degree()
\end{sageexample}

In order to test a specific point $(r,d)$, the data array must contain at least $(r+1)(d+2)$ terms. 
If it has more terms, the guess becomes more trustworthy, but also the computation time increases.
By setting the option \verb|ensure| to a positive integer~$e$, the user can request that only 
such points $(r,d)$ should be tested for which the data array contains at least $e$ more terms than
needed. This increases the reliability. By setting the option \verb|cut| to a positive integer~$c$,
the user requests that for testing a point $(r,d)$, the method should take into account at most $c$
more terms than needed. If the data array contains more terms, superfluous ones are ignored in the
interested of better performance. We must always have $0\leq e\leq c\leq \infty$. The default setting
is $e=0$, $c=\infty$.

Finally, there is an option \verb|infolevel| which, when set to a positive integer, causes the method
to print some output about the progress of the computation. Higher values lead to more detailed output.

\section{Concluding Remarks}

...
